Higher-Order unification for free!

### 2024-05-09 paper/main.tex
add node nb

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling problematic sub terms on
the side.
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It is worth mentioning that we replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable. % by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goals but is consistently faster than
the Coq one starting from goals made of XXX nodes.
\todo{D: how many?}
Moreover our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Finally, our encoding is sufficiently shallow to: 1)
allow to lift forms of static analysis for
the meta language, such as determinacy, also to the object
language; 2) take advantage of future improvements to the
logic-programming engine of Elpi, such as tabled search.

>>>>

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling problematic sub terms on
the side.
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It is worth mentioning that we replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable. % by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goals but is consistently faster than
the Coq one starting from goals made of $32$ nodes.
Moreover our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Finally, our encoding is sufficiently shallow to: 1)
allow to lift forms of static analysis for
the meta language, such as determinacy, also to the object
language; 2) take advantage of future improvements to the
logic-programming engine of Elpi, such as tabled search.
